---
title: "Conversion to and from R data"
---

One of the key goals with extendr, is to provide a framework that allows you to write Rust functions, that interact with R, without having to know the intricacies within R internals, or even R's C-facilities. However, this is unavoidable if one wishes to understand why the extendr-api is the way it is.

Thus, for introducing extendr, we shall mention facts about R internals, but these are not necessary to keep in mind going forward.

```{r, echo=FALSE}
# to enable extendr/extendrsrc chunks
library(rextendr)
# names(knitr::knit_engines$get())
```

A fundamental data-type in R is the 32-bit integer, `int` in C, and `i32` in Rust. Passing that type around is essential, and straight forward:

```{extendrsrc}
#[extendr]
fn ultimate_answer() -> i32 {
    return 42_i32;
}
```

And now this function is available within your R-session, as the output is `r ultimate_answer()`.

Also, another fundamental data-type in R is `numeric` / `f64`, which we can also pass back and forth uninhibitated, e.g.

```{extendrsrc}
#[extendr]
fn return_tau() -> f64 {
    std::f64::consts::TAU
} 
```

where $\tau := 2\pi =$ $`r return_tau()`$.

However, passing data from R to Rust must be done with a bit of care: In R, representing a true integer in literal form requires using `L` after the literal. 

```{extendrsrc}
#[extendr]
fn bit_left_shift_once(number: i32) -> i32 {
    number << 1
}
```

This function supposedly is a clever way to multiply by two, however passing `bit_left_shift_once(21.1)` results in

```{r, error=TRUE, echo=FALSE}
bit_left_shift_once(21.1)
```
where `bit_left_shift_once(21)` is `r bit_left_shift_once(21L)`, as expected.

R also has the concept of missing numbers, `NA` encoded within its data-model. However `i32`/`f64` do not natively have a representation for `NA` e.g.

```{r, error=TRUE}
bit_left_shift_once(NA_integer_)
bit_left_shift_once(NA_real_)
bit_left_shift_once(NA)
```

Instead, we have to rely on extendr's scalar variants of R types, `Rint` / `Rfloat` to encompass the notion of `NA` in our functions:

```{extendrsrc}
#[extendr]
fn double_me(value: Rint) -> Rint {
    if value.is_na() {
        Rint::na()
    } else {
        (value.inner() << 1).into()
    }
}
```

which means, we can now handle missing values in the arguments

```{r, error=FALSE}
double_me(NA_integer_)
double_me(NA_real_)
double_me(NA)
```

One may notice here that `NA_real_` was accepted even for an `Rint`. The reason
for this, is when you specify a type without `&`/`&mut`, the value is coerced
in a similar way, as R coerces values. In order to have strict type-checking
during run-time, use `&` / `&mut`, as


```{extendrsrc}
#[extendr]
fn wrong_input(value: Rint) -> Rint {
    value.clone()
}
```

```{r, error=TRUE}
wrong_input(NA_integer_)
wrong_input(NA_real_)
wrong_input(21.0)
wrong_input(21L)
```

Here, only the last literal is a true `Rint`. 

## Vectors

Most data in R are vectors. Scalar values are in fact 1-sized vectors, and
even lists are defined by a vector-type. A vector type in Rust is `Vec`. A
`Vec` has a type-information, length, and capacity. This means, that if necessary,
we may expand any given `Vec`-data to contain more values, and only when capacity
is exceeded, will there be a reallocation.

Naively, we may define a function like so

```{extendrsrc}
#[extendr]
fn repeat_us(mut values: Vec<i32>) -> Vec<i32> {
    assert_eq!(values.capacity(), values.len(), "must have zero capacity left");
    values[0] = 100;
    values.push(55);
    values
}
```

```{r, error=TRUE}
x <- c(1L, 2L, 33L)
repeat_us(x)
```

Even if the argument is `mut Vec<_>`, what happens is that the R vector gets
converted to a Rust owned type, and it is that type that we can modify, and augment, with syncing to the original data.

Of course, a slice e.g. `&[i32]` / `&mut [i32]` could be used instead, and this allows us to modify the original data, i.e.

```{extendrsrc}
#[extendr]
fn zero_middle_element(mut values: Robj) {
    use extendr_api::AsTypedSlice;
    let values: &mut [i32] = values.as_typed_slice_mut().unwrap();
    let len = values.len();
    let middle = len / 2;
    values[middle] = 0;
}
```

```{r}
x <- c(100L, 200L, 300L)
zero_middle_element(x)
x
```

This is great! If we wanted to insert an `NA` in the middle, we would have had to operate on `&mut [Rint]` instead. 

A slice is a representation of a sequence of elements that are part of a larger collection. Since they represent only part of a collection (vector, in this case), we cannot add new elements to this. To do so,  we have to rely on extendr provided types, that provide a `Vec`-like API to R's vector-types. These are the `Integers`, `Logicals`, `Doubles`, and `Strings` types.

## Strings are special

The R runtime performs [string interning](https://en.wikipedia.org/wiki/String_interning) to
all of its string elements. This means, that whenever R encounters a new string,
it adds it to its internal string intern pool. 

::: {.callout-tip }
A string intern pool can be thought of as a container that stores all distinct
strings, and then provides a lightweight reference counted variable back to it.
An example of such a string interner is the [`lasso`](https://crates.io/crates/lasso) crate.
:::

Let's look at a concrete example:

```{extendrsrc}
#[extendr]
fn hello_world() -> &'static str {
    "Hello world!"
}
```

```{r}
.Internal(inspect(hello_world()))
```

Then, any time R encounters `"Hello world!"`, it retrieves it from the pool, rather
than re-instantiate it

```{r}
.Internal(inspect("Hello world!"))
```

The `STRSXP` is different, due to R's clone semantics, but the underlying
string `CHARSXP` is the same. Thus, equality is determined if two strings
have the same pointer, rather than if they have the same bytes.

Therefore, `extendr` does not provide mutable access to an R string, because it breaks
the assumption that all strings are the immutable.